Posterior summarisation in Bayesian phylogenetics using Tracer 1.7

### 2017-12-19 manuscript/tracer17.tex
Checks whether statistics are still being calculated.

Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) \citep{rannala1996probability, mau1999bayesian} flourishes as a popular approach to uncover the evolutionary relationships among taxa, such as genes, genomes, individuals or species.
MCMC approaches generate samples of model parameter values - including the phylogenetic tree - drawn from their posterior distribution given molecular sequence data and a selection of evolutionary models.
Visualising, tabulating and marginalising these samples is critical for approximating the posterior quantities of interest that one reports as the outcome of a Bayesian phylogenetic analysis.
To facilitate this task, we have developed the \tracer\ (version 1.7) software package to
process MCMC trace files containing parameter samples and to interactively explore the high-dimensional posterior distribution.
\tracer\ works automatically with sample output from BEAST 1 \citep{drummond2012bayesian}, BEAST 2 \citep{bouckaert2014beast2},  LAMARC \citep{kuhner2006lamarc},  Migrate \citep{beerli2006comparison}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes} and possibly other MCMC programs from other domains.

@@

\tracer\ offers a solution of visualising conditional posterior distributions as well.
A typical use case involves Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters influence the likelihood function only when a specific model is selected by a random indicator function.
Under BSSVS, a posterior estimate of the parameter should only sample values from states where the indicator equals one.
Discrete phylogeographic analyses frequently employ BSSVS due to the potentially large amount of transition rates that need to be estimated \citep{Lemey2009}, but this is also relevant when employing model averaging approaches such as for relaxed molecular clocks for example \citep{Li2012}.
%
Finally, \tracer\ provides demographic reconstruction resulting in a graphical plot, often applied to reconstruct epidemic dynamics.
%Depending on the actual demographic model used in the analysis, an additional file containing samples from the posterior tree distribution will be required.
%The currently
Available models are constant size, exponential and logistic growth,
%\citep{drummond2002estimating},
%Bayesian  skyline \citep{drummond2005bayesian},
%Bayesian skyride \citep{minin2008smooth} and
%Bayesian skygrid \citep{gill2012improving}.
and the non-parametric Bayesian skline, skyride and skygrid.
%Not sure if the statement below is correct ...
%All of these demographic models are available in BEAST 1 \citep{drummond2012bayesian} and BEAST 2 \citep{bouckaert2014beast2}.
%Besides these, ``Lineages Through Time'' (?) is also available to plot the quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?

@@

\begin{figure}
\subfigure[TangHuLu chart]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integerjointmarginal.pdf}}
\subfigure[Marginal density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integermarginaldensity.pdf}} \\
\subfigure[Scatter plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-correlation.pdf}}
\subfigure[Correlation plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-joint-real.pdf}} \\
%\subfigure[Box-and-whisker plot]{\includegraphics[width = .465\textwidth, height = 3cm]{./figures/rabv-catreal.pdf}}
\caption{Joint marginal and marginal density plots of: (a) two integer variables through a TangHuLu chart; (b) frequency plots the marginal density of multiple integer or categorical variables; (c) classic scatter or correlation plot for two continuous variables; (d) large correlation matrices for multiple ($> 2$) continuous variables.
%; (e) box-and-whisker plots for the joint marginal distribution of a continuous and an integer or categorical variable.
}
\label{fig:4tabs}
\end{figure}

@@

Phylogeographic BSSVS inference includes parameters of both integer (number of non-zero transition rates) and categorical (host or location-state) trace type.
%While a single parameter of such types can easily be visualised using a frequency plot,
In \tracer, we have developed a TangHuLu chart to visualise the joint probability distribution between two integer or categorical traces using coloured bubbles (see Figure \ref{fig:4tabs}:a).
Here, circle size is proportional to the joint probability, %the blue coloured circle is in the credible set of given a probability threshold default to 0.95, the red is in the set not in the credible set.
with a coloured tile background if this probability reaches a nominal threshold to enhance visibility.
%The coloured background can also help to show the area of the credible set and non-credible set.
%Multiple integer parameters can also be visualised using
Marginal density plots (see Figure \ref{fig:4tabs}:b) can also display multiple integer parameters, each with unique colour scales.
With approximately equal numbers of transition rates, both figures suggest a similar host and location trait model complexity.
%Both figures show a clear overlap in the number of non-zero transition rates in the discrete trait models for host and location, as identified using a BSSVS procedure.
\tracer\ also provides
%The visualisation options for integer or categorical parameters are complemented with
 popular visualisations for continuous parameters, including scatter plots (see Figure \ref{fig:4tabs}:c) for two parameters,
 and recent extensions for correlations between $\ge\hspace{-0.32em}2$ continuous parameters (Figure \ref{fig:4tabs};d; \citet{Murdoch}).
 %, which indicates a strong positive correlation between the clock rates of both discrete trait models.
%We also provide an extension of such a scatter plot for multiple (i.e. more than two) continuous variables (see Figure \ref{fig:4tabs}:d; \citet{Murdoch}), which makes use of coloured ellipses to indicate the strength of correlation between each pair of continuous variables.
Colour gradients indicate the strength and direction of the correlation, from red (strong negative-correlation) to blue (strong positive-correlation).
%with a colour gradient from white to dark blue indicating increasingly strong correlation and a colour gradient from white to dark red indicating increasingly strong anti-correlation.
The shape of ellipse drawn re-enforces the strength of correlation, with no correlation appearing as a circle and perfect (anti-)correlation as a line.
%Finally, in order to display the joint marginal distribution between a continuous variable and an integer or categorical variable, we make use of box-and-whisker plots.
%Using such a plot (Figure \ref{fig:4tabs}:e), we can infer that the range of the non-zero transition rates can depend on the ancestral location state inferred at the root of the underlying phylogeny.

>>>>

Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) \citep{rannala1996probability, mau1999bayesian} flourishes as a popular approach to uncover the evolutionary relationships among taxa, such as genes, genomes, individuals or species.
MCMC approaches generate samples of model parameter values - including the phylogenetic tree - drawn from their posterior distribution given molecular sequence data and a selection of evolutionary models.
Visualising, tabulating and marginalising these samples is critical for approximating the posterior quantities of interest that one reports as the outcome of a Bayesian phylogenetic analysis.
To facilitate this task, we have developed the \tracer\ (version 1.7) software package to
process MCMC trace files containing parameter samples and to interactively explore the high-dimensional posterior distribution.
\tracer\ works automatically with sample output from BEAST \citep{drummond2012bayesian}, BEAST2 \citep{bouckaert2014beast2},  LAMARC \citep{kuhner2006lamarc},  Migrate \citep{beerli2006comparison}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes} and possibly other MCMC programs from other domains.

@@

\tracer\ offers a solution of visualising conditional posterior distributions as well.
A typical use case involves Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters influence the likelihood function only when a specific model is selected by a random indicator function.
Under BSSVS, a posterior estimate of the parameter should only sample values from states where the indicator equals one.
Discrete phylogeographic analyses frequently employ BSSVS due to the potentially large amount of transition rates that need to be estimated \citep{Lemey2009}, but this is also relevant when employing model averaging approaches such as for relaxed molecular clocks for example \citep{Li2012}.
%
Finally, \tracer\ provides demographic reconstruction resulting in a graphical plot, often applied to reconstruct epidemic dynamics.
%Depending on the actual demographic model used in the analysis, an additional file containing samples from the posterior tree distribution will be required.
%The currently
Available models are constant size, exponential and logistic growth,
%\citep{drummond2002estimating},
%Bayesian  skyline \citep{drummond2005bayesian},
%Bayesian skyride \citep{minin2008smooth} and
%Bayesian skygrid \citep{gill2012improving}.
and the non-parametric Bayesian skline, skyride and skygrid.
%Not sure if the statement below is correct ...
%All of these demographic models are available in BEAST \citep{drummond2012bayesian} and BEAST2 \citep{bouckaert2014beast2}.
%Besides these, ``Lineages Through Time'' (?) is also available to plot the quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?

@@

\begin{figure}
\subfigure[Bubble chart]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integerjointmarginal.pdf}}
\subfigure[Marginal density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integermarginaldensity.pdf}} \\
\subfigure[Scatter plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-correlation.pdf}}
\subfigure[Correlation plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-joint-real.pdf}} \\
%\subfigure[Box-and-whisker plot]{\includegraphics[width = .465\textwidth, height = 3cm]{./figures/rabv-catreal.pdf}}
\caption{Joint marginal and marginal density plots of: (a) two integer variables through a bubble chart; (b) frequency plots the marginal density of multiple integer or categorical variables; (c) classic scatter or correlation plot for two continuous variables; (d) large correlation matrices for multiple ($> 2$) continuous variables.
%; (e) box-and-whisker plots for the joint marginal distribution of a continuous and an integer or categorical variable.
}
\label{fig:4tabs}
\end{figure}

@@

Phylogeographic BSSVS inference includes parameters of both integer (number of non-zero transition rates) and categorical (host or location-state) trace type.
%While a single parameter of such types can easily be visualised using a frequency plot,
In \tracer, we have developed a bubble chart to visualise the joint probability distribution between two integer or categorical traces using coloured bubbles (see Figure \ref{fig:4tabs}:a).
Here, circle size is proportional to the joint probability, %the blue coloured circle is in the credible set of given a probability threshold default to 0.95, the red is in the set not in the credible set.
with a coloured tile background if this probability reaches a nominal threshold to enhance visibility.
%The coloured background can also help to show the area of the credible set and non-credible set.
%Multiple integer parameters can also be visualised using
Marginal density plots (see Figure \ref{fig:4tabs}:b) can also display multiple integer parameters, each with unique colour scales.
With approximately equal numbers of transition rates, both figures suggest a similar host and location trait model complexity.
%Both figures show a clear overlap in the number of non-zero transition rates in the discrete trait models for host and location, as identified using a BSSVS procedure.
\tracer\ also provides
%The visualisation options for integer or categorical parameters are complemented with
 popular visualisations for continuous parameters, including scatter plots (see Figure \ref{fig:4tabs}:c) for two parameters,
 and recent extensions for correlations between $\ge\hspace{-0.32em}2$ continuous parameters (Figure \ref{fig:4tabs};d; \citet{Murdoch}).
 %, which indicates a strong positive correlation between the clock rates of both discrete trait models.
%We also provide an extension of such a scatter plot for multiple (i.e. more than two) continuous variables (see Figure \ref{fig:4tabs}:d; \citet{Murdoch}), which makes use of coloured ellipses to indicate the strength of correlation between each pair of continuous variables.
Colour gradients indicate the strength and direction of the correlation, from red (strong negative-correlation) to blue (strong positive-correlation).
%with a colour gradient from white to dark blue indicating increasingly strong correlation and a colour gradient from white to dark red indicating increasingly strong anti-correlation.
The shape of ellipse drawn re-enforces the strength of correlation, with no correlation appearing as a circle and perfect (anti-)correlation as a line.
%Finally, in order to display the joint marginal distribution between a continuous variable and an integer or categorical variable, we make use of box-and-whisker plots.
%Using such a plot (Figure \ref{fig:4tabs}:e), we can infer that the range of the non-zero transition rates can depend on the ancestral location state inferred at the root of the underlying phylogeny.
